---
title: "Ninja Warrior"
output: 
  html_notebook: 
    number_sections: yes
    toc: yes
    fig_width: 25
    fig_height: 10
editor_options: 
  chunk_output_type: inline
---

 Read in with tidytuesdayR package 
 Install from CRAN via: install.packages("tidytuesdayR")
 This loads the readme and all the datasets for the week of interest


```{r}
require(tidytuesdayR)
require(tidyverse)
library(tidytext)
```


Read Data


```{r}
tuesdata <- tidytuesdayR::tt_load('2020-12-15')
ninja_warrior <- tuesdata$ninja_warrior

ninja_warrior

```

Check data

```{r}
glimpse(ninja_warrior)
```
```{r}

```

National Finals are held in Las Vegas in US and Sasuke(Japan)
```{r}
ggplot(by_loc_stage_obst, aes(x = "", fill = factor(round_stage))) +
        geom_bar(width = 1) +
        theme(axis.line = element_blank(), 
              plot.title = element_text(hjust=0.5)) + 
        labs(fill="Stage",
             x=NULL, 
             y=NULL, 
             title="Pie Chart of Stages at locations", 
             caption="Source: Ninja Warroir") +
            coord_polar(theta = "y", start=0) +
            facet_wrap(~location,drop=TRUE)
```


```{r}
ninja_warrior %>% 
  filter(str_detect(round_stage,"National Finals")) %>% 
  ggplot(aes(x=location,fill=round_stage))+
  geom_bar(position="dodge")
```

Semi-Finals are held at Venice only

```{r}
ninja_warrior %>% 
  filter(round_stage == "Semi-Finals") %>% 
  ggplot(aes(x=location,fill=round_stage))+
  geom_bar(position="dodge")
```


```{r}
obstacles <- ninja_warrior %>% 
              group_by(obstacle_name) %>% 
                summarise(n=n()) %>% 
                  arrange(desc(n))

```

The highest used obstacles (>30) and at which stage they are used the most


```{r}
ninja_warrior %>% 
  filter(obstacle_name %in% c('Warped Wall','Salmon Ladder','Quintuple Steps')) %>% 
    ggplot(aes(round_stage,fill=obstacle_name)) +
      geom_bar(position='dodge') +
        coord_flip()

```

What are stage names?

```{r}
unique(ninja_warrior$round_stage)
```



How are obstacles used at each stage ?


Qualifying

```{r}
ninja_warrior %>% 
  group_by(round_stage,obstacle_name ) %>% 
    filter(round_stage=="Qualifying")  %>% 
      mutate(count=n()) %>% 
      ggplot(aes(obstacle_name,count)) +
        geom_col(aes(fill = obstacle_name), color = NA) +
           labs(x = "", y = "no of times obstacle is used") +
             coord_polar()  +
              guides(fill = FALSE) 
```
Qualifying (Regional/City)
```{r}
ninja_warrior %>% 
  group_by(round_stage,obstacle_name ) %>% 
    filter(round_stage=="Qualifying (Regional/City)")  %>% 
      mutate(count=n()) %>% 
      ggplot(aes(obstacle_name,count)) +
        geom_col(aes(fill = obstacle_name), color = NA) +
           labs(x = "", y = "no of times obstacle is used") +
             coord_polar()  +
              guides(fill = FALSE)
```



Semi-Finals

```{r}
ninja_warrior %>% 
  group_by(round_stage,obstacle_name ) %>% 
    filter(round_stage=="Semi-Finals")  %>% 
      mutate(count=n()) %>% 
      ggplot(aes(obstacle_name,count)) +
        geom_col(aes(fill = obstacle_name), color = NA) +
           labs(x = "", y = "no of times obstacle is used") +
             coord_polar()  +
              guides(fill = FALSE)
```
```{r}

obstacles_by_round <-  ninja_warrior %>% 
                          group_by(round_stage,obstacle_name ) %>% 
                            
                              mutate(round_stage = as.factor(round_stage)) %>% 
                              count(obstacle_name,sort=TRUE) %>% 
                                ungroup()


obstacles_by_round %>% 
                  group_by(round_stage) %>% 
                    top_n(10) %>% 
                      ungroup %>%
                        mutate(round_stage = as.factor(round_stage),
                               obstacle_name = reorder_within(obstacle_name, n,round_stage)) %>%
                                ggplot(aes(obstacle_name, n, fill = round_stage)) +
                                  geom_col(show.legend = FALSE) +
                                      facet_wrap(~round_stage, scales = "free_y") +
                                        coord_flip() +
                                          scale_x_reordered() +
                                          scale_y_continuous(expand = c(0,0)) +
                                            labs(y = "Number of obstacles per stage",
                                                 x = NULL,
                                                 title = "Highest obstacles per stage?"
                                                 )
```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
