---
title: "A ggplot2 Tutorial for Beautiful Plotting in R"
output:
  html_document:
    df_print: paged
layout: post
showtoc: yes
description: An extensive tutorial containing a general introduction to ggplot2 as
  well as many examples how to modify a ggplot, step by step. It covers several topics
  such as different chart types, themes, design choices, plot combinations, and modification
  of axes, labels, and legends, custom fonts, interactive charts and many more.
tags:
- DataViz
- tutorial
- R
- tidyverse
- ggplot2
editor_options:
  chunk_output_type: console
---

```{r set options, include=FALSE}
options(width = 700)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width = "100%",
                      fig.showtext = TRUE, retina = 2)
```
## Table of Content {#toc}

* [Preparation](#prep)
* [The Dataset](#data)
* [The `{ggplot2}` Package](#ggplot)
* [A Default ggplot](#default)
* [Working with Axes](#axes)
* [Working with Titles](#titles)
* [Working with Legends](#legends)
* [Working with Backgrounds & Grid Lines](#style)
* [Working with Margins](#margins)
* [Working with Multi-Panel Plots](#panels)
* [Working with Colors](#colors)
* [Working with Themes](#themes)
* [Working with Lines](#lines)
* [Working with Text](#text)
* [Working with Coordinates](#coords)
* [Working with Chart Types](#charts)
* [Working with Ribbons (AUC, CI, etc.)](#ribbons)
* [Working with Smoothings](#smooths)
* [Working with Interactive Plots](#interactive)
* [Remarks, Tipps & Resources](#tipps)




## Preparation {#prep}


````{r install-packages, eval=FALSE}
# install CRAN packages
install.packages(c("tidyverse", "colorspace", "corrr",  "cowplot",
                   "ggdark", "ggforce", "ggrepel", "ggridges", "ggsci",
                   "ggtext", "ggthemes", "grid", "gridExtra", "patchwork",
                   "rcartocolor", "scico", "showtext", "shiny",
                   "plotly", "highcharter", "echarts4r"))

# install from GitHub since not on CRAN
install.packages(devtools)
devtools::install_github("JohnCoene/charter")

library(tidyverse)

```



## The Dataset {#data}

We are using data from the *National Morbidity and Mortality Air Pollution Study* (NMMAPS). To make the plots manageable we are limiting the data to Chicago and 1997–2000. For more detail on this data set, consult Roger Peng's book [Statistical Methods in Environmental Epidemiology with R](http://www.springer.com/de/book/9780387781662).
You can download the data we are using during this tutorial [here](https://github.com/Z3tt/R-Tutorials/blob/master/ggplot2/chicago-nmmaps.csv) (but you don't have to).

We can import the data into our R session for example with `read_csv()` from the `{readr}` package. To access the data later, we are storing it in a variable called `chic` by using the *assignment arrow* `<-`.

```{r}
chic <- readr::read_csv("https://raw.githubusercontent.com/Z3tt/R-Tutorials/master/ggplot2/chicago-nmmaps.csv")


```
**The `::` is called *namespace* and can be used to access a function without loading the package. Here, you could also run `library(readr)` first and `chic <- read_csv(...)` afterwards.**

```{r  data-inspection}
tibble::glimpse(chic)

head(chic)
```

## The `{ggplot2}` Package {#ggplot}

> `ggplot2` is a system for declaratively creating graphics, based on [The Grammar of Graphics](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=as_li_ss_tl?ie=UTF8&qid=1477928463&sr=8-1&keywords=the+grammar+of+graphics&linkCode=sl1&tag=ggplot2-20&linkId=f0130e557161b83fbe97ba0e9175c431). You provide the data, tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
A ggplot is built up from a few basic elements:

1.	**Data**:  
The raw data that you want to plot.
2.	**Geometries** `geom_`:  
The geometric shapes that will represent the data.
3.	**Aesthetics** `aes()`:  
Aesthetics of the geometric and statistical objects, such as position, color, size, shape, and transparency
4.	**Scales** `scale_`:  
Maps between the data and the aesthetic dimensions, such as data range to plot width or factor values to colors.
5.	**Statistical transformations** `stat_`:  
Statistical summaries of the data, such as quantiles, fitted curves, and sums.
6.	**Coordinate system** `coord_`:  
The transformation used for mapping data coordinates into the plane of the data rectangle.
7.	**Facets** `facet_`:  
The arrangement of the data into a grid of plots.
8.	**Visual themes** `theme()`:  
The overall visual defaults of a plot, such as background, grids, axes, default typeface, sizes and colors.


💡 **The number of elements may vary depending on how you group them and whom you ask.**

## A Default ggplot {#default}

First, to be able to use the functionality of `{ggplot2}` we have to load the package (which we can also load via the [tidyverse package collection](https://www.tidyverse.org/)):

```{r library}
#library(ggplot2)
library(tidyverse)
```

The syntax of `{ggplot2}` is different from base R. In accordance with the basic elements, a default ggplot needs three things that you have to specify: the *data*, *aesthetics*, and a *geometry*. We always start to define a plotting object by calling `ggplot(data = df)` which just tells `{ggplot2}` that we are going to work with that data. In most cases, you might want to plot two variables—one on the x and one on the y axis. These are *positional aesthetics* and thus we add `aes(x = var1, y = var2)` to the `ggplot()` call (yes, the `aes()` stands for aesthetics). However, there are also cases where one has to specify one or even three or more variables.

💡 **We specify the data *outside* `aes()` and add the variables that ggplot maps the aesthetics to *inside* `aes()`.**

Here, we map the variable `date` to the x position and the variable `temp` to the y position. Later, we will also map variables to all kind of other aesthetics such as color, size, and shape.

```{r ggplot}
(g <- ggplot(chic, aes(x = date, y = temp)))
```

Hm, only a panel is created when running this. Why? This is because `{ggplot2}` does not know *how* we want to plot that data—we still need to provide a geometry!

`ggplot2` allows you to store the current `ggobject` in a variable of your choice by assigning it to a variable, in our case called `g`. You can extend this `ggobject` later by adding other layers, either all at once or by assigning it to the same or another variable.

💡 **By using parentheses while assigning an object, the object will be printed immediately (instead of writing `g <- ggplot(...)` and then `g` we simply write `(g <- ggplot(...))`).**

There are many, many different geometries (called *geoms* because each function usually starts with `geom_`) one can add to a ggplot by default (see [here](https://ggplot2.tidyverse.org/reference/) for a full list) and even more provided by extension packages (see [here](https://exts.ggplot2.tidyverse.org/) for a collection of extension packages). Let's tell `{ggplot2}` which style we want to use, for example by adding `geom_point()` to create a scatter plot:

```{r ggplot-default}
g + geom_point()
```

Nice! But this data could be also visualized as a line plot (not optimal, but people do things like this all the time). So we simply add `geom_line()` instead and voilá:

```{r ggplot-default-line}
g + geom_line()
```

One can also combine several geometric layers—and this is where the magic and fun starts!

```{r ggplot-default-line-point}
g + geom_line() + geom_point()
```

That's it for now about geometries. No worries, we are going to learn several plot types at [a later point](#charts).

##### Change Properties of Geometries

Within the `geom_*` command, you already can manipulate visual aesthetics such as the color, shape, and size of your points. Let's turn all points to large fire-red diamonds!

```{r ggplot-default-col-size-shape}
g + geom_point(color = "firebrick", shape = "diamond", size = 2)

```

💡 **`{ggplot2}` understands both `color` and `colour` as well as the short version `col`.**

<details>
  <summary>&#128129; <i>You can use preset colors (here is a [full list](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)) or [hex color codes](https://www.techopedia.com/definition/29788/color-hex-code), both in quotes, and even RGB/RGBA colors by using the `rgb()` function. Expand to see example.</i></summary>


```{r ggplot-default-col-size-hex-rgb, eval=FALSE}
g + geom_point(color = "#b22222", shape = "diamond", size = 2)
g + geom_point(color = rgb(178, 34, 34, maxColorValue = 255), shape = "diamond", size = 2)
```

```{r ggplot-default-col-size-hex-rgb-plot, echo=FALSE, purl=FALSE}
library(patchwork)
a <- g + geom_point(color = "#b22222", shape = "diamond", size = 2)
b <- g + geom_point(color = rgb(178, 34, 34, maxColorValue = 255), shape = "diamond", size = 2)
a + b
```

</details>

Each geom comes with its own properties (called *arguments*) and the same argument may result in a different change depending on the geom you are using.

```{r ggplot-default-line_col-size-shape}
g + geom_point(color = "firebrick", shape = "diamond", size = 2) +
    geom_line(color = "firebrick", linetype = "dotted", size = .3)
```

##### Replace the default `ggplot2` theme

And to illustrate some more of ggplot's versatility, let's get rid of the grayish default `{ggplot2}` look by setting a different built-in theme, e.g. `theme_bw()`—by calling `theme_set()` all following plots will have the same black'n'white theme. The red points look way better now!

```{r}
theme_set(theme_bw())

g+geom_point(color="firebrick")

```

You can find more on how to use built-in themes and how to customize themes in the section ["Working with Themes"](#themes). From the next chapter on, we will also use the `theme()` function to customize particular elements of the theme.

💡 **`theme()` is an essential command to manually modify all kinds of theme elements (texts, rectangles, and lines).**

To see which details of a ggplot theme can be modified have a look [here](https://ggplot2.tidyverse.org/reference/theme.html)—and take some time, this is a looong list.

[↑ Jump back to Table of Content.](#toc)<br><br>

## Working with Axes {#axes}

Let's add some well-written labels to the axes. For this, we add `labs()` providing a character string for each label we want to change (here `x` and `y`):

```{r axis-labs}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)")
```

<details>
  <summary>&#128129; <i>You can also add each axis title via `xlab()` and `ylab()`. Expand to see example.</i></summary>
```{r axis-labs-2}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  xlab("Year") +
  ylab("Temperature (°F)")
```
</details>


Usually you can also specify symbols by simply adding the symbol itself (here "°") but the code below also allows to add not only symbols but e.g. superscripts:

```{r axis-labs-expression}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = expression(paste("Temperature (", degree ~ F, ")"^"(Hey, why should we use metric units?!)")))
```


##### Increase Space between Axis and Axis Titles

`theme()` is an essential command to modify particular theme elements (texts and titles, boxes, symbols, backgrounds, ...). We are going to use them a lot! For now, we are going to modify text elements. We can change the properties of all or particular text elements (here axis titles) by overwriting the default `element_text()` within the `theme()` call:

```{r labs-move-away-vjust}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title.x = element_text(vjust = 0, size = 15),
        axis.title.y = element_text(vjust = 2, size = 15))
```

`vjust` refers to the vertical alignment, which usually ranges between 0 and 1 but you can also specify values outside that range. Note that even though we move the axis title on the y axis horizontally, we need to specify `vjust` (which is correct form the label's perspective). You can also change the distance by specifying the margin of both text elements:

```{r labs-move-away-margin}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title.x = element_text(margin = margin(t = 10), size = 15),
        axis.title.y = element_text(margin = margin(r = 10), size = 15))
```

The labels `t` and `r` within the `margin()` object refer to *top* and *right*, respectively. You can also specify the four margins as `margin(t, r, b, l)`. Note that we now have to change the right margin to modify the space on the y axis, not the bottom margin.

💡 **A good way to remember the order of the margin sides is "*t*-*r*-oub-*l*-*e*".**

##### Change Aesthetics of Axis Titles

Again, we use the `theme()` function and modify the element `axis.title` and/or the subordinated elements `axis.title.x` and `axis.title.y`. Within the `element_text()` we can for example overwrite the defaults for `size`, `color`, and `face`:

```{r labs-color-axes-1}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title = element_text(size = 15, color = "firebrick",
                                  face = "italic"))
```

The `face` argument can be used to make the font `bold` or `italic` or even `bold.italic`.

```{r labs-color-axes-2}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title.x = element_text(color = "sienna", size = 15),
        axis.title.y = element_text(color = "orangered", size = 15))
```

<details>
  <summary>&#128129; <i>You could also use a combination of `axis.title` and `axis.title.y`, since `axis.title.x` inherits the values from `axis.title`. Expand to see example.</i></summary>
```{r labs-color-axes-3}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title = element_text(color = "sienna", size = 15),
        axis.title.y = element_text(color = "orangered", size = 15))
```
</details>

One can modify some properties for both axis titles and other only for one or properties for each on its own:

```{r labs-color-axes-4}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.title = element_text(color = "sienna", size = 15, face = "bold"),
        axis.title.y = element_text(face = "bold.italic"))

```


##### Change Aesthetics of Axis Text

Similarly, you can also change the appearance of the axis text (here *the numbers*) by using `axis.text` and/or the subordinated elements `axis.text.x` and `axis.text.y`:

```{r labs-color-axes-text}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.text = element_text(color = "dodgerblue", size = 12),
        axis.text.x = element_text(face = "italic"))
```



##### Rotate Axis Text

Specifying an `angle` allows you to rotate any text elements. With `hjust` and `vjust` you can adjust the position of the text afterwards horizontally (0 = left, 1 = right) and vertically (0 = top, 1 = bottom):

```{r axis-text}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1, size = 12))
```



##### Remove Axis Text & Ticks

There may be rarely a reason to do so—but this is how it works:

```{r axis-no-labs}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```

I introduced three theme elements—text, lines, and rectangles—but actually there is one more: `element_blank()` which removes the element (and thus is not considered an official element).

💡 **If you want to get rid of a theme element, the element is always `element_blank()`.**

##### Remove Axis Titles

We could again use `theme_blank()` but it is way simpler to just remove the label in the `labs()` (or `xlab()`) call:

```{r axis-no-title}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = NULL, y = "")
```

💡 **Note that `NULL` removes the element (similarly to `element_blank()`) while empty quotes `""` will keep the spacing for the axis title and simply print nothing.**

##### Limit Axis Range

Sometimes you want to <s>zoom into</s> take a closer look at some range of your data. You can do this without subsetting your data:

```{r axis-limit}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  ylim(c(0, 50))
```

Alternatively you can use `scale_y_continuous(limits = c(0, 50))` or `coord_cartesian(ylim = c(0, 50))`. The former removes all data points outside the range while the second adjusts the visible area and is similar to `ylim(c(0, 50))`. You may wonder: *So in the end both result in the same.* But not really, there is an important difference—compare the two following plots:

```{r axis-limit-comp, echo=FALSE, purl=FALSE}
p1 <- ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  scale_y_continuous(limits = c(0, 50)) +
  ggtitle("scale_y_continuous(limits) or ylim()")
p2 <- ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (°F)") +
  coord_cartesian(ylim = c(0, 50)) +
  ggtitle("coord_cartesian(ylim)")
library(patchwork)
p1 + p2
```

You might have spotted that on the left there is some empty buffer around your y limits while on the right points are plotted right up to the border and even beyond. This perfectly illustrates the subsetting (left) versus the zooming (right). To show why this is important let's have a look at a different chart type, a box plot:

```{r axis-limit-comp-box, echo=FALSE, purl=FALSE}
p1 <- ggplot(chic, aes(x = factor(year), y = temp)) +
  geom_boxplot() +
  labs(x = "Year", y = "Temperature (°F)") +
  scale_y_continuous(limits = c(0, 50)) +
  ggtitle("scale_y_continuous(limits) or ylim()")
p2 <- ggplot(chic, aes(x = factor(year), y = temp)) +
  geom_boxplot() +
  labs(x = "Year", y = "Temperature (°F)") +
  coord_cartesian(ylim = c(0, 50)) +
  ggtitle("coord_cartesian(ylim)")
p1 + p2
```

Um. Because `scale_x|y_continuous()` subsets the data first, we get completely different (and wrong, at least if in the case this was not your aim) estimates for the box plots! I hope you don't have to go back to your old scripts now and check if you *maybe* have manipulated your data while plotting and did report wrong summary stats in your report, paper or thesis...


##### Force Plot to Start at Origin

Related to that, you can force R to plot the graph starting at the origin:

```{r origin}
library(tidyverse)
chic_high <- dplyr::filter(chic, temp > 25, o3 > 20)
ggplot(chic_high, aes(x = temp, y = o3)) +
  geom_point(color = "darkcyan") +
  labs(x = "Temperature higher than 25°F",
       y = "Ozone higher than 20 ppb") +
  expand_limits(x = 0, y = 0)
```

<details>
  <summary>&#128129; <i>Using `coord_cartesian(xlim = c(0, NA), ylim = c(0, NA))` will lead to the same result. Expand to see example.</i></summary>
```{r origin-coord}
library(tidyverse)
chic_high <- dplyr::filter(chic, temp > 25, o3 > 20)
ggplot(chic_high, aes(x = temp, y = o3)) +
  geom_point(color = "darkcyan") +
  labs(x = "Temperature higher than 25°F",
       y = "Ozone higher than 20 ppb") +
  coord_cartesian(xlim = c(0, NA), ylim = c(0, NA))
```
</details>

But we can also force it to *literally* start at the origin!

```{r origin-force}
ggplot(chic_high, aes(x = temp, y = o3)) +
  geom_point(color = "darkcyan") +
  labs(x = "Temperature higher than 25°F",
       y = "Ozone higher than 20 ppb") +
  expand_limits(x = 0, y = 0) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off")
```

💡 **The argument `clip = "off"` in any coordinate system, always starting with `coord_*`, allows to draw outside of the panel area.**

Here, I call it to make sure that the tick marks at `c(0, 0)` are not cut. See the [Twitter thread by Claus Wilke](https://twitter.com/clauswilke/status/991542952802619392?lang=en) for more details.

```
##### Axes with Same Scaling

For demonstrating purposes, let's plot temperature against temperature with some random noise. The `coord_equal()` is a coordinate system with a specified ratio representing the number of units on the y-axis equivalent to one unit on the x-axis. The default, `ratio = 1`, ensures that one unit on the x-axis is the same length as one unit on the y-axis:

```{r axes-equal}
ggplot(chic, aes(x = temp, y = temp + rnorm(nrow(chic), sd = 20))) +
  geom_point(color = "sienna") +
  labs(x = "Temperature (°F)", y = "Temperature (°F) + random noise") +
  xlim(c(0, 100)) + ylim(c(0, 150)) +
  coord_fixed()
```

Ratios higher than one make units on the y axis longer than units on the x-axis, and vice versa:

```{r axes-fixed-2, fig.height=3}
ggplot(chic, aes(x = temp, y = temp + rnorm(nrow(chic), sd = 20))) +
  geom_point(color = "sienna") +
  labs(x = "Temperature (°F)", y = "Temperature (°F) + random noise") +
  xlim(c(0, 100)) + ylim(c(0, 150)) +
  coord_fixed(ratio = 1/5)
  ```
  
```{r}

ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = NULL) +
  scale_y_continuous(label = function(x) {paste(x, "Degrees Fahrenheit")})
```

##### Use a Function to Alter Labels

Sometimes it is handy to alter your labels a little, perhaps adding units or percent signs without adding them to your data. You can use a function in this case:

```{r labs-alt}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = NULL) +
  scale_y_continuous(label = function(x) {return(paste(x, "Degrees Fahrenheit"))})  
```

[↑ Jump back to Table of Content.](#toc)<br><br>
  

