---
title: "TidyTemplate"
date: 2021-02-11
output: html_output
editor_options: 
  chunk_output_type: console
---

# TidyTuesday

Join the R4DS Online Learning Community in the weekly #TidyTuesday event!
Every week we post a raw dataset, a chart or article related to that dataset, and ask you to explore the data.
While the dataset will be “tamed”, it will not always be tidy! As such you might need to apply various R for Data Science techniques to wrangle the data into a true tidy format.
The goal of TidyTuesday is to apply your R skills, get feedback, explore other’s work, and connect with the greater #RStats community!
As such we encourage everyone of all skills to participate!

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytuesdayR)
library(scales)
theme_set(theme_light())
```

# Load the weekly Data

Download the weekly data and make available in the `tt` object.

```{r Load}

tt <- tt_load("2021-02-09")

```


# Readme

Take a look at the readme for the weekly data to get insight on the dataset.
This includes a data dictionary, source, and a link to an article on the data.

```{r Readme, eval = interactive()}

tt

```


# Glimpse Data

Take an initial look at the format of the data available.

```{r Glimpse}

tt %>% 
  map(glimpse)

```


```{r}

#Let's make one graph of each dataset
tt$lifetime_earn %>% 
  ggplot(aes(lifetime_earn,race,fill=gender))+
  geom_col(position="dodge") +
  scale_x_continuous(labels=dollar)



plot_by_race <- function(data,column,labels=dollar,...){
   last_year <-data %>%
              group_by(race) %>% 
             slice_max(year,n=1)
  
   data %>% 
    mutate(race= fct_reorder(race,-{{column}})) %>% 
    ggplot(aes(year,{{column}},color=race,...)) +
    geom_line() +
     geom_text(aes(label=race,color=race),hjust=0,data=last_year,nudge_x = 0.3)+
    expand_limits(x=2022,y=0) +
    scale_y_continuous(labels = labels) +
    labs(
      x="Year",
      color="Race"
    )+
    theme(
      legend.position = "none",
      panel.grid = element_blank()
    )
}



tt$student_debt %>% 
  plot_by_race(loan_debt_pct,percent)+
  labs(y= "% of families with student loan debt")

tt$student_debt %>% 
  plot_by_race(loan_debt,dollar) +
  labs(y= "Avg student loan debt")

tt$retirement %>% 
    plot_by_race(retirement,dollar)+
    labs(y="Average family liquid retirement savings normalized to 2016 dollars")


tt$home_owner %>% 
  plot_by_race(home_owner_pct,labels=percent) +
  labs(y="Home ownership percentage")

tt$race_wealth %>% 
  plot_by_race(wealth_family,labels=dollar) +
  facet_wrap(~type,scales = "free_y")+
  expand_limits(x=2025)+
  labs(y ="Family wealth (2016 dollars)")

tt$income_time %>% 
  mutate(percentile=fct_reorder(percentile,-income_family)) %>% 
  ggplot(aes(year,income_family,color=percentile)) +
  geom_line() +
  expand_limits(y=0)+
  scale_y_continuous(labels=dollar)+
  labs(x ="Year",y="Family Income")
  

tt$income_time %>% 
  pivot_wider(names_from = percentile,values_from = income_family) %>% 
  ggplot(aes(year,`50th`,ymin=`10th`,ymax=`90th`))+
  geom_ribbon(alpha=0.25) +
  geom_line(size=1) +
  scale_y_continuous(labels=dollar)+
  labs(x="Year",y="Family-level income with(median with 10th and 90th percentiles")

tt$income_limits %>% 
  
  filter(dollar_type == "2019 Dollars",
         !str_detect(race,"or in Combination")) %>%
  filter(income_quintile == "Top 5%") %>% 
  distinct(race,year,income_quintile,.keep_all = TRUE) %>% 
  plot_by_race(income_dollars,labels=dollar) +
  expand_limits(x=2028)+
  labs(y="Top 5% income (2019 dollars)")


tt$income_limits %>% 
  filter(dollar_type == "2019 Dollars",
         !str_detect(race,"or in Combination")) %>%
  distinct(race,year,income_quintile,.keep_all = TRUE) %>% 
  pivot_wider(names_from = income_quintile,values_from=income_dollars) %>% 
  mutate(race=fct_reorder(race,-Fourth)) %>% 
  ggplot(aes(year,ymin=Lowest,ymax=Fourth,fill=race))+
  geom_ribbon(alpha=0.25) +
  scale_y_continuous(labels=dollar)+
  labs(x="Year",y="20th-80th quantiles")

library(plotly)
g <- tt$income_limits %>% 
        filter(dollar_type == "2019 Dollars",
               !str_detect(race,"or in Combination")) %>%
        distinct(race,year,income_quintile,.keep_all = TRUE) %>% 
        mutate(income_quintile=fct_reorder(income_quintile,-income_dollars)) %>% 
        ggplot(aes(year,income_dollars,color=income_quintile))+
        geom_line() +
        scale_y_continuous(labels=dollar)+
        labs(x="Year",y="Income Quintile",color="") +
        facet_wrap(~race)
  
ggplotly(g)


```

```{r}
tt$income_aggregate %>% 
  group_by(year,race) %>% 
  filter(income_quintile  != "Top 5%") %>% 
  summarise(sum(income_share))


tt$income_aggregate %>% 
  group_by(year,race) %>% 
   filter(income_quintile  != "Top 5%",
          !str_detect(race,"or in Combination")) %>% 
  mutate(income_share = income_share/100,
         income_quintile = fct_inorder(income_quintile)) %>% 
  ggplot(aes(year,income_share,fill=income_quintile)) +
  geom_area() +
  facet_wrap(~race)+
  scale_y_continuous(labels=percent)+
  labs(y="% share of income",
       fill="Income Quintile")

```



























